Detailed structure of phospholipid bilayers with negatively charged lipids and salt concentrations from molecular dynamics simulations including the effects of electronic polarization

### 2019-05-17 Manuscript/manuscript.tex
Added gamma label to POPS chemfig and removed associated todo
Moved one paragraph from counterion affinity to structural agreement (previous) section

\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/Order-parameters_form-factors_exp-L17-ECC-lipids.pdf}
  \includegraphics[width=\figwidth]{../img/POPSstructure.pdf} 
\hfill
  \caption{\label{simVSexpNOions_POPS} 
    \textbf{Top:} X-ray scattering form factors from simulations with Lipid17/Dang \citep{lipid17-future, dang2006} and 
    ECC-POPS/ECC-ions \cite{martinek17, Pluhackova2016} compared with experiments~\citep{kucerka14} at 298~K. 
    \textbf{Middle:} Order parameters of POPS head group, glycerol backbone and acyl chains  
    from the same simulations 
    compared with experiments at 298~K. \citep{NMRlipidsIV}
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPS. 
    \textbf{Bottom:} The chemical structure of POPS and the labeling of the carbon segments. 
  }  
\todo{Add label $\gamma$ to the figure -- it is used in Fig with the dihedral angle, 
and it can be also used in the description of COO- vector. }
\end{figure}

@@

Despite the success in simulating bilayer dimensions and acyl chain conformations,
lipid force fields have typically problems in capturing the correct glycerol backbone and
headgroup order parameters and conformations \cite{botan15,ollila16,NMRlipidsIV}.
This is the case also for Lipid17 POPS simulations, where the headgroup and glycerol backbone
order parameters are quite far from experimental values with both Dang (Fig. \ref{simVSexpNOions_POPS})
and {\AA}qvist \cite{NMRlipidsIV} ions. The values from ECC-POPS simulation are closer,
but not in full agreement with experiments (Fig.~\ref{simVSexpNOions_POPS}).
Especially, the experimentally observed more negative values of the $\beta$-carbon
and the larger forking (difference of order parameters between hydrogens attached to the
same carbon \cite{ollila16}) of the $\alpha$-carbon in POPS than in POPC \cite{NMRlipidsIV}
are not reproduced by neither of the models here. 
These differences are previously related
to a more rigid structure of PS headgroup and 
the stiffness of the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral
plotted in Fig.~\ref{fig:dihedral}. 
Interestingely, this dihedral angle is more flexible in ECC-POPS compared to Lipid17 
suggesting that its rigidity is overestimated in Lipid17 due to the lack of electronic polarizability,
while the same dihedral parameters yield too small forking in ECC-POPS.
Although the ECC-POPS model reproduces the lipid bilayer dimensions and
acyl chain conformations in a good agreement with experiments, 
there is a room for improvements especially in the glycerol backbone and headgroup region,
e.g. in the parameters of the associated dihedral angles. 
\cite{botan15, ollila16, Pluhackova2016, NMRlipidsIV}.

@@

\subsection{Counterion binding affinity and interactions between PC and PS head groups}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_purePOPS_ecclipids-lipid17.pdf}
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_5PC-1PS_ecclipids-lipid17.pdf}
  \caption{\label{fig:POPS-counterions-dens}
    Number density profiles of \ce{K^{+}} and \ce{Na^{+}} counterions along the membrane normal axis
    in ECC-lipids (solid lines) and Lipid17/Dang (dashed lines) simulations of pure POPS (left) or POPC:POPS (5:1, right) bilayers.  
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
}
\end{figure} 
Binding affinity of monovalent ions to lipid bilayers depends strongly on
force field parameters \cite{catte16,NMRlipidsIV},
which is the case also here as the sodium counterion binding affinity to
POPS lipid bilayer is stronger in Lipid17/ Dang simulation than in
ECC-POPS simulation (Fig. \ref{fig:POPS-counterions-dens}).
The cation binding affinities to lipid bilayers in simulations have been previously evaluated
against experiments using the order parameters of $\alpha$ and $\beta$ carbons in the
phosphatidylcholine (PC) lipid headgroup (see Fig.~\ref{fig:chemstruct_pc_ps} for the labeling) \cite{catte16,melcr18,NMRlipidsIV},
which decrease proportionally to the bound charge according to the ''electrometer concept'' \citep{seelig87}.
Althought the response of PS lipid headgroup order parameters to the bound charge is also systematic,
it is less well undestood and the addition of cations may cause phase transitions in negatively charged
bilayers \cite{feigenson86,mattai89,roux91,roux90}. Therefore, the cation binding affinity to bilayers with
PS lipids has been evaluated measuring the PC headgroup order parameters from POPC:POPS (5:1)
mixtures\cite{roux90,NMRlipidsIV}.
Following our previous work \cite{NMRlipidsIV}, we evaluate the counterion binding affinity
to lipid bilayer in simulations containing PS using two different approaches.

% The situation is more complicated for lipid bilayers containing
%PS lipids, for which there is no ion-free reference state due to the presence of counterions.

%Unlinke in pure PC bilayers,
%Interpreting the changes of the head group order parameters of 
%mixed PC:PS lipid bilayers 
%from both simulations and experiments
%is complicated by the presence of counterions. \cite{NMRlipidsIV}

The head group and acyl chain order parameters within ECC-lipids
are in general in a good agreement with the experimental values 
as shown in Fig. \ref{simVSexpNOions_POPS}. 
The acyl chain order parameters in particular agree with the experimental data within error bars.
The order parameters of the head groups are at an accuracy comparable to 
other currently available classical models of lipids \citep{botan15, catte16, Pluhackova2016, nmrlipids_proj4}. 
In particular, ECC-POPS is closer to the experimental order parameters 
than Lipid17 with either Dang \cite{smith94,chang1999,dang2006} 
or ff99 ions \citet{aqvist90}.

%This was demonstrated for a bound positive charge in simulations with POPC and 
%cationic surfactants or aquaeous cations employing the ECC-POPC model \cite{melcr18}.
%In this work, we will use the electrometer concept also for a bound negative charge 
%using PS lipids as negatively charged surfactants. \cite{roux90}

%is the
%is not generally agreed on between simulations 
%\citep{bockmann03,sachs04,berkowitz06,cordomi09}
%and experiments 
%\citep{cevc90,tocanne90,hauser76,herbette84,uhrikova08}.
%We attempted to find a model, 
%which would successfully interpret the experiments on binding of \ce{Na^+}
%to neutral and negatively charged phospholipid bilayers \citep{akutsu81, roux90} 
%in our works \citep{catte16, NMRlipidsIV}.
%After testing the currently available force fields for phospholipids,
%we concluded
%that the interactions are generally overestimated in magnitude in almost all models 
%but Lipid14 (PC) \citep{dickson14}, resp. Lipid17 (PS) \citep{lipid17-future}. 
%This model yields a semi-quantitative agreement with the experimentally measured small changes of the order parameters \cite{catte16}
%when used with the model of ions by \citet{aqvist90}. 
%However, when used with a more accurate model of ions by \citet{Pluharova2014, martinek17},
%the model overestimates the binding affinity of \ce{Na^+}
%measured with lipid electrometer concept. \citep{melcr18}
%In total, these results suggest that improvements 
%in the lipid parameters are required for more accurate interactions even with monovalent cations. 
%\citep{catte16, melcr18, NMRlipidsIV}

>>>>

\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/Order-parameters_form-factors_exp-L17-ECC-lipids.pdf}
  \includegraphics[width=\figwidth]{../img/POPSstructure.pdf} 
\hfill
  \caption{\label{simVSexpNOions_POPS} 
    \textbf{Top:} X-ray scattering form factors from simulations with Lipid17/Dang \citep{lipid17-future, dang2006} and 
    ECC-POPS/ECC-ions \cite{martinek17, Pluhackova2016} compared with experiments~\citep{kucerka14} at 298~K. 
    \textbf{Middle:} Order parameters of POPS head group, glycerol backbone and acyl chains  
    from the same simulations 
    compared with experiments at 298~K. \citep{NMRlipidsIV}
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPS. 
    \textbf{Bottom:} The chemical structure of POPS and the labeling of the carbon segments. 
  }  
\end{figure}

@@

Despite the success in simulating bilayer dimensions and acyl chain conformations,
lipid force fields have typically problems in capturing the correct glycerol backbone and
headgroup order parameters and conformations \cite{botan15,ollila16,NMRlipidsIV}.
This is the case also for Lipid17 POPS simulations, where the headgroup and glycerol backbone
order parameters are quite far from experimental values with both Dang (Fig. \ref{simVSexpNOions_POPS})
and {\AA}qvist \cite{NMRlipidsIV} ions. The values from ECC-POPS simulation are closer,
but not in full agreement with experiments (Fig.~\ref{simVSexpNOions_POPS}).
Especially, the experimentally observed more negative values of the $\beta$-carbon
and the larger forking (difference of order parameters between hydrogens attached to the
same carbon \cite{ollila16}) of the $\alpha$-carbon in POPS than in POPC \cite{NMRlipidsIV}
are not reproduced by neither of the models here. 
These differences are previously related
to a more rigid structure of PS headgroup and 
the stiffness of the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral
plotted in Fig.~\ref{fig:dihedral}. 
Interestingely, this dihedral angle is more flexible in ECC-POPS compared to Lipid17 
suggesting that its rigidity is overestimated in Lipid17 due to the lack of electronic polarizability,
while the same dihedral parameters yield too small forking in ECC-POPS.

Overall,
the head group and acyl chain order parameters within ECC-lipids
are in general in a good agreement with the experimental values 
as shown in Fig. \ref{simVSexpNOions_POPS}. 
The acyl chain order parameters in particular agree with the experimental data within error bars.
The order parameters of the head groups are at an accuracy comparable to 
other currently available classical models of lipids \citep{botan15, catte16, Pluhackova2016, nmrlipids_proj4}. 
In particular, ECC-POPS is closer to the experimental order parameters 
than Lipid17 with either Dang \cite{smith94,chang1999,dang2006} 
or ff99 ions \citet{aqvist90}. 
Although the ECC-POPS model reproduces the lipid bilayer dimensions and
acyl chain conformations in a good agreement with experiments, 
there is a room for improvements especially in the glycerol backbone and headgroup region,
e.g. in the parameters of the associated dihedral angles. 
\cite{botan15, ollila16, Pluhackova2016, NMRlipidsIV}.

@@

\subsection{Binding of counterions to POPS and POPC and interactions between their head groups}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_purePOPS_ecclipids-lipid17.pdf}
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_5PC-1PS_ecclipids-lipid17.pdf}
  \caption{\label{fig:POPS-counterions-dens}
    Number density profiles of \ce{K^{+}} and \ce{Na^{+}} counterions along the membrane normal axis
    in ECC-lipids (solid lines) and Lipid17/Dang (dashed lines) simulations of pure POPS (left) or POPC:POPS (5:1, right) bilayers.  
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
}
\end{figure} 
In simulations, binding affinity of monovalent ions to lipid bilayers depends strongly on
force field parameters \cite{catte16,NMRlipidsIV}.
For instance, the affinity of sodium counterions to a POPS bilayer 
is stronger in Lipid17/Dang simulations 
than in simulations with ECC-POPS (Fig.~\ref{fig:POPS-counterions-dens}).
The cation binding affinities to lipid bilayers in simulations have been previously evaluated
against experiments using the order parameters of $\alpha$ and $\beta$ carbons in the
phosphatidylcholine (PC) lipid headgroup (see Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC} for the labeling) \cite{catte16,melcr18,NMRlipidsIV},
which decrease proportionally to the bound charge according to the ''electrometer concept'' \citep{seelig87}.
Although the response of PS lipid headgroup order parameters to the bound charge is also systematic,
it is less well undestood and the addition of cations may cause phase transitions in negatively charged
bilayers \cite{feigenson86,mattai89,roux91,roux90}. Therefore, the cation binding affinity to bilayers with
PS lipids has been evaluated measuring the PC headgroup order parameters from POPC:POPS (5:1)
mixtures\cite{roux90,NMRlipidsIV}.
Following our previous work \cite{NMRlipidsIV}, we evaluate the counterion binding affinity
to a lipid bilayer containing PS in simulations using two approaches.
